---
title: "697D-Final"
author: "Elaona Lemoto, Katherine (Yang) Liu"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(magrittr)
library(ggplot2)
library(dplyr)
library(ggforce)
library(tidyr)
library(readr)
library(purrr)
library(tidyverse)
```


##Introduction

Social determinants of health are defined as environmental factors, such as, where someone lives, where they go to school, work, and so forth, and how it affects a whole set of health outcomes and their overall quality of life. For instance, for folks who live in a middle class neighborhood might have higher access to grocery stores or farmers markets and a recreational facility, comparable to lower income neighborhoods that are found to have more fast food restaurants, less available grocery stores and more supercenters, and lack of recreational facilites. Social determinants of health is a growing field as more and more people are attempting to better understand how environmental factors impact one's health.

For this project in particular, we aim to study the social determinants of food security. We would like to look at how environmental factors like access to stores, proximity to stores, assistance for food by State, etc., impacts rates of food insecurity. What we currently understand about food insecurity is that it is not an isolated act. While income plays a part in one's ability to access food, there are several other variables that impact one’s security in having access to food. What the FDA proposed through the collection of these data is that different parts of a household's environment can impact their access to food. Therefore, the purpose of this project is to utilize the varying factors of one’s household in the geological, social, and economic sense from the variables outlined above, and uncover or magnify any measured impacts or explanation of a State’s percentage of households with food insecurity, with low food insecurity, and child food insecurity. 


##Data

In order to study this even further, we found data from the United States Department of Agriculture under their Economic Research Service. They published a dataset with over 275 variables, uniquely identified by FIPS county codes. For each county, they collected categories of data like Access, Health, Local Foods, and Price Taxes etc. Within these categories, this dataset contains several variables with comprehensive data on counts of households that have or do not have access to supermarkets, counts of store availability and whether it is within a walking distance of homes in that county, counts of fast food restaurant availability near home, counts of WIC redemptions, and number of local foods and farms, as well as variables concerning  socioeconomic status. Also included in this dataset were variables concerning food insecurity: household food insecurity, household low food insecurity, and child food insecurity, however, accounted for at the State and not FIPS county level. 


##Research Questions

Using the data from the FDA, we are interested in measuring the varying geographic distance to resources, social net programs and assistance, and/or economic costs or barriers have any correlation with the aggregate percentage of households with food insecurity, households with low food security, and child food insecurity in at the State level.

A secondary question we would like to explore whether or not there are certain variables or factors that capture much of the variation we might see in our food insecurity variables.As mentioned above, the dataset created by the ERS department of the FDA has placed all of their variables in categories which we will use to create a factor analysis of food
insecurity. 

An additional possible question we would like to explore is whether a factor of socioeconomic status holds a significant impact on food insecurity. Redlining in the U.S. was a systematic discriminatory practice that divested from predominantly black and brown neighborhoods, therefore, we should expect to see a significance in socioeconomic demographics and food insecurity in the U.S.


##Method


###Cleaning Data

Before taking the weighted average of each of our predictor variables of interst, all of our variables were continuous on a 0 to 100 scale indicating a proportion of their respective county.To prepare our data for this study, we took the weight of each county in each state by its population size relative to the overall population,  and multiplied each of our predictor variables by that weight before taking the average value. 

- PCT_LACCESS_POP15 = percent of population with low access to a stores in 2015
- PCT_LACCESS_HHNV15 = percent of households with no car and low access to stores in 2015
- PCT_LACCESS_SNAP15 = percent of SNAP households with low access to stores in 2015
- PCT_LACCESS_CHILD15 = percent of children with low access to store in 2015
- PCT_NHWHITE10 = percent of neighborhood that is white 
- PCT_NHBLACK10 = percent of neighborhood that is black
- PCT_HISP10 = percent of neighborhood that is hispanic
- PCT_NHASIAN10 = percent of neighborhood that is asian
- PCT_NHNA10 = percent of neighborhood that is native american
- PCT_NHPI10 = percent of neighborhood that is native hawaiian and pacific islander 
- GROCPTH16 = percent of grocery stores per thousand people people in 2016
- SUPERCPTH16 = percent of super centers per thousand people people
- PCT_SNAP17 = percent of SNAP redemtions in a county
- WICSPTH16 = percent of WIC authorized stores per thousand people
- FFRSPTH16 = percent of Fast food restaurants per thousand people in 2016
- FSRSPTH16 = percent of full service restaurants per thousand people in 2016
- PCT_REDUCED_LUNCH15 =  percent of students eligible for reduced-price lunch in 2015
- PCT_OBESE_ADULTS17 = percent of adults with obesity in 2017
- RECFACPTH16 = percent of Recreation & fitness facilities per thousand people in 2016


```{r}
#Bring in data and do some cleaning
(df <- read_csv("final_data.csv"))
df<-df[-1,]
df<-df%>% 
  select(-FOODINSEC_12_14, -VLFOODSEC_12_14,  -CH_VLFOODSEC_14_17, -SNAPSPTH17)
n <- nrow(df)

#Response
hist(df$FOODINSEC_15_17)
hist(df$VLFOODSEC_15_17)
hist(df$CH_FOODINSEC_14_17)

#Predictors
hist(df$PCT_LACCESS_POP15)
hist(df$PCT_LACCESS_HHNV15)
hist(df$PCT_LACCESS_SNAP15)
hist(df$PCT_LACCESS_CHILD15)
hist(df$PCT_NHASIAN10)
hist(df$PCT_NHWHITE10)
hist(df$PCT_NHBLACK10)
hist(df$PCT_HISP10)
hist(df$PCT_NHNA10)
hist(df$PCT_NHPI10)
hist(df$GROCPTH16)
hist(df$PCT_REDUCED_LUNCH15)
hist(df$PCT_OBESE_ADULTS17)
hist(df$PCT_SNAP17)
hist(df$WICSPTH16)
hist(df$)
summary(df)


```



###Multivariable Application

In order to conduct this study, we will begin by conducting a multivariate
regression model with all of our predictor variables of interest and our three explained variables regarding food insecurity. Our predictor variables are indicated above. We will be interpeting the multivariate linear model function calculations in R. 
Included in this method will be simultaneous confidence intervals and predictor variables.

From here we will utilize the bootstrapping method for evaluating point estimates from R-squared. This will be done by creating bootstrap sample estimates for R-squared and then confidence intervals for R-squared based on the distribution of bootstrap distributions we simulate. Our second method with bootstrapping will be conducted by calculating the likelihood ratio test statistic of one of or more of our predictor variables and comparing it to the F distribution and the approximate chi-squared distribution. 

Another method we will be using from our course is conducting a factor analysis of our categories. This will implement our two last research questions regarding the categories of variables that the FDA’s dataset created as well as measuring whether certain factors do a better job of explaining the variation in food insecurity. 


##Results


```{r}
# using the lm()
res <- lm(cbind(FOODINSEC_15_17, CH_FOODINSEC_14_17, VLFOODSEC_15_17) ~ ., data = df)
summary(res)

res2 <- lm(cbind(FOODINSEC_15_17, CH_FOODINSEC_14_17, VLFOODSEC_15_17) ~ PCT_NHBLACK10+ PCT_HISP10+GROCPTH16 + SUPERCPTH16 + WICSPTH16 + FFRPTH16 + PCT_REDUCED_LUNCH15+RECFACPTH16, data = df)
summary(res2)

# Residuals:
head(resid(res))
# Fitted values:
head(fitted(res))
# Residual standard error:
sigma(res)
round(vcov(res), 2)
```

Above we have our multivariate linear model exploring the potential impacts of our variates of interest and food insecurity. Looking at our variable, Food Insecurity from 2015 to 2017, we find very strong relationships with Fast Food Restaurants for every thousand people and percent of county with free or reduced lunch in 2015. Secondary relationships we found were with WIC-accepted Stores per thousand people and Recreational Centers per thousand people. We found these results to be interesting as assistance in the sense of reducing the cost of food and indicators of food insecure areas like number of Fast food restaurants and larger number of stores accepting state food support, had significant coefficents. Looking at the magnitude of reduced lunch, it seems as though there is a potential benefit to having more assistance in reducing food costs. When it comes to the indicators of food insecurity, the presence of fast food restaurants seems to be very signicant and looking at the magnitude, a much stronger potential indicator of food insecure areas. The same idea translates to our signifance to percentage of Wic accepted stores. Although we would think that with more WIC-accepted stores we would see a negative magnitude due to a reduced cost to food, but in terms of our data and linear model, could be viewed as an indicator instead. 

Moving towards our second multivariate linear model, we are looking at drawing relationships between our predictors and Child Food Insecurity. Comparable to our model with Food Insecurity, we found less coefficents from our predictors to hold signifcance in this model. We found percentage of neighborhood that is Black to have a signicantly negative relationship with Child Food Insecurity, percent of grocery stores per thousand residents to have a postive and very large magnitude, and lastly, percent of population that receives reduced lunch to have a signicantly negative relationship. We found these results to be pretty surprising considering that one of our research questions sought to look at the relationship between racial demographics and food insecurity. In this case our model tested and found that with higher percentage of black communites, we on average are likely to see less child food insecurity, which is the opposite of the result we were expecting to see and we aren't quite sure how to interpret that with a real world context. The other result of a positive relationship between percentage of grocery stores and food insecurity was also surprising. One one end, having more grocery stores could signal a less food insecure state simply because there is access to more perishable goods comparable to what a supercenter, fast food, or full service restaurant could provide, but on the other hand it doesn't indicate whether residents can still access the grocery stores due to potential higher prices. 


Lastly, here are the results for our response variable: Households with Very Low Food security. Again, we found pretty signicant relationships between our coefficients for percent of county enrolled in free or reduced lunch and percent of WIC-accepted grocery stores per thousand people. An added interest was the associated signifcance from the coefficent for percentage of neighborhood that is Hispanic. In this case, our estimated coefficent was negative meaning with higher percentage of hispanic residents, we should expect a smaller amount of food insecurity. Again here we see a different result than we were expecting in terms of race and food security which was very surpising. We also saw a bit of significance from our coeffecients on percent of recreational faciliites per thousand people and percent of fast food restaurants per thousand residents.  Similar to our results with our Food Insecurity response variable, we see that there are competing relationships where the coefficient for reduced lunch and recreation facilities provide a negative relationship to food insecurity, whereas the presence of WIC supported stores and fast food restaurants potentially have a postive coefficent in regards to households with very low food security. 


Now that we ran our multivariate linear model with all of our predictor variables, we also ran a restricted model including all of the variables that were signicant in our previous model. While we did find some signifnicance in our coefficents for WIC-accepted grocery stores, our R-squared value decresed significantly from our full model to the adjusted model, so we can't say that the model is more efficient when restricted. 


Overall, what we found were competing relationships between what we would consider indcators of more food insecure areas and assistance for reducing the costs of food. What we thought we would also see was a separate relationship with racial demographics and food insecurity, however, that was not something we could find with this current data. 


```{r}
alpha <- 0.05

Y <- as.matrix(df %>% select(20:22))
m <- ncol(Y) # number of responses
Z <- as.matrix(cbind(const = 1, df %>% select(1:19)))
r <- ncol(Z) - 1  # number of predictors (excluding constantn
b <- solve(t(Z) %*% Z) %*% t(Z) %*% Y

pred <- Z %*% b
resid <- Y - pred
estSigma <- t(resid) %*% resid / n

# new observation for predictors 
z0 <- 
  matrix(
    c(1, 25, 5, 1.3,  0.5, 34, 39, 9,3,0.2, 0.04, 0.24, 0.006, 1, 0.09, 0.6, 15, 
      18, 4.7, 25.6),
    ncol=1, 
    dimnames = list(c("const", "PCT_LACCESS_POP15", "PCT_LACCESS_HHNV15", "PCT_LACCESS_SNAP15", "LACCESS_CHILD15", "PCT_NHWHITE10", "PCT_NHBLACK10", "PCT_HISP10", "PCT_NHASIAN10", "PCT_NHNA10", "PCT_NHPI10", "GROCPTH16", "SUPERCPTH16", "WICSPTH16", "FFRPTH16", "FSRPTH16", "PCT_SNAP17", "PCT_REDUCED_LUNCH15", "PCT_OBESE_ADULTS17", "RECFACPTH16"), NULL)
  )

#choose z0 that is meaningful 

# center
cntr <- t(b) %*% z0

# Prediction interval:
fctr_pred <- 
  sqrt(
    (1 + t(z0) %*% solve(t(Z) %*% Z,tol = 1e-19) %*% z0) * 
      (m * (n - r - 1) * qf(alpha, m, n - r - m, lower.tail = FALSE) / (n - r - m)))

A <- diag(3)
for (i in 1:3){
  a <- A[ ,i]
  half_CI <- (sqrt(t(a) %*% (estSigma * n / (n - r - 1)) %*% a)) * fctr_pred
  pred_CI_Y <- c(t(a) %*%  cntr - half_CI, t(a) %*%  cntr + half_CI)
  cat(paste0("95% simultaneous prediction interval for Y0",i, ":  (",pred_CI_Y[1],", ",pred_CI_Y[2], ")\n"))
}

# Confidence interval: 
fctr <- 
  sqrt(
    (t(z0) %*% solve(t(Z) %*% Z) %*% z0) * 
      (m * (n - r - 1) * qf(alpha, m, n - r - m, lower.tail = FALSE) / (n - r - m)))

for (i in 1:3){
  a <- A[ ,i]
  half_CI <- (sqrt(t(a) %*% (estSigma * n / (n - r - 1)) %*% a)) * fctr
  pred_CI_Y <- c(t(a) %*%  cntr - half_CI, t(a) %*%  cntr + half_CI)
  cat(paste0("95% simultaneous confidence interval for Y0",i, ":  (",pred_CI_Y[1],", ",pred_CI_Y[2], ")\n"))
}
```

Above we have our simultaneous confidence interval and simultaneous prediction intervals.

# Factor Analysis

```{r}
#Food Insecurity
df_f1 <- df %>% select(20:22)
R <- cor(df_f1)
eig <- eigen(R)
(Ltilde_m1 <- sqrt(eig$values[1]) * eig$vectors[ , 1, drop = F])

psi_tilde_m1 <- diag(R - Ltilde_m1 %*% t(Ltilde_m1))
Psi_tilde_m1 <- diag(psi_tilde_m1)
residual_matrix_m1 <- R - Ltilde_m1 %*% t(Ltilde_m1) - Psi_tilde_m1
(eig$values[1] / sum(eig$values))

Ltilde_m1 %*% t(Ltilde_m1)

#Access

df_f2 <- df %>% select(1:4)
R2<- cor(df_f2)
eig2 <- eigen(R2)
(Ltilde_m2 <- sqrt(eig2$values[1]) * eig2$vectors[ , 1, drop = F])


psi_tilde_m2 <- diag(R2 - Ltilde_m2 %*% t(Ltilde_m2))
Psi_tilde_m2 <- diag(psi_tilde_m2)
residual_matrix_m2 <- R2 - Ltilde_m2 %*% t(Ltilde_m2) - Psi_tilde_m2
(eig2$values[1] / sum(eig2$values))

#Demographics
df_f3 <- df %>% select(5:10)
R3 <- cor(df_f3)
eig3 <- eigen(R3)
(Ltilde_m3 <- sqrt(eig3$values[1]) * eig3$vectors[ , 1, drop = F])

psi_tilde_m3 <- diag(R3 - Ltilde_m3 %*% t(Ltilde_m3))
Psi_tilde_m3 <- diag(psi_tilde_m3)
residual_matrix_m3 <- R3 - Ltilde_m3 %*% t(Ltilde_m3) - Psi_tilde_m3
(eig3$values[1] / sum(eig3$values))

#StoreAccess
df_f4 <- df %>% select(11:16)
R4 <- cor(df_f4)
eig4 <- eigen(R4)
(Ltilde_m4 <- sqrt(eig4$values[1]) * eig4$vectors[ , 1, drop = F])

psi_tilde_m4 <- diag(R4 - Ltilde_m4 %*% t(Ltilde_m4))
Psi_tilde_m4 <- diag(psi_tilde_m4)
residual_matrix_m4 <- R4 - Ltilde_m4 %*% t(Ltilde_m4) - Psi_tilde_m4
(eig4$values[1] / sum(eig4$values))


```

Above we have the factor analysis using the Principal Component Method and our different factors of interest. In this case it was Racial Demographics, Health, Store Accesss, Access, and Food Insecurity.

Looking at Food Insecurity, we see that the total variance for this factors is somewhat explained by one factor, however, it is not very strong. Looking at the residual matrix, the non-diagonal elements seem to be small or at least close to zero. Therefore, the factor for Food Insecurity seems sufficient.

Evaluating Store Access with the variables: Grocery store per thousand people, Super Center per thousand people, WICS-accepted stores per thousand people, Fast food restaurants per thousand people, Full-service restaurants and percent of SNAP redemptions in a county. Evaluting its residual matrix, all of the surrounding elements are very small which tells us that Store Access is a sufficient factor.


Another similar factor was Access. Looking at our variables concering geographical access to a store near households, 93 percent of its variation can be explained by a single factor. Also evaluating the residual matrix calculated from our eigenvalue and vectors, we can see that Access is a very sufficient factor variable. 

And lastly, one of our interests was to look at the potential realtionship between racial demographics. Here we can see if our demographic variables would be sufficient for a single factor. Evaluating the single factors 


From here we will utilize the bootstrapping method for evaluating point estimates from R-squared. This will be done by creating bootstrap sample estimates for R-squared and then confidence intervals for R-squared based on the distribution of bootstrap distributions we simulate. Our second method with bootstrapping will be conducted by calculating the likelihood ratio test statistic of one of or more of our predictor variables and comparing it to the F distribution and the approximate chi-squared distribution.

#Bootstrapping
##Likelihood Ratio Test
```{r}
variables_to_test <-  c("PCT_LACCESS_POP15","PCT_LACCESS_HHNV15","PCT_LACCESS_SNAP15","PCT_LACCESS_CHILD15","PCT_NHWHITE10", "PCT_NHASIAN10","PCT_NHNA10","PCT_NHPI10", "FSRPTH16","PCT_SNAP17", "PCT_OBESE_ADULTS17")

variables_to_keep <- setdiff(colnames(Z), variables_to_test)
q <- length(variables_to_keep) - 1 # Not counting the intercept
grp1 <- rownames(b) %in% variables_to_keep
grp2 <- rownames(b) %in% variables_to_test

estSigma <- t(resid) %*% resid / n

b_grp1 <- solve(t(Z[ , grp1]) %*% Z[ , grp1]) %*% t(Z[ , grp1]) %*% Y
pred_grp1 <- Z[ , grp1] %*% b_grp1
resid_grp1 <- Y - pred_grp1

estSigma1 <- t(resid_grp1) %*% resid_grp1 / n

LR_test_statistic <- -n * (log(det(estSigma)) - log(det(estSigma1)))

# Approximate likelihood ratio test statistic:
approx_LR_test_statistic <- 
  -(n - r - 1 - 0.5 * (m - r + q + 1)) * (log(det(estSigma)) - log(det(estSigma1)))

# Reject?
ifelse(approx_LR_test_statistic > qchisq(1-alpha, m * (r - q)), "Reject H0", "Don't reject H0")

W <- det(estSigma) / det(estSigma1)

h <- length(variables_to_test)  
# h = q' = r - q; see corrected slide 25 in notes for Chapter 7

d1 <- (n - r - 1) - 0.5 * (m - h + 1)
d2 <- (m * h - 2)/4
d3 <- m^2 + h^2 - 5
d3 <- if (d3 > 0) { 
  sqrt(((m * h)^2 - 4) / d3)
} else { 1 }

F_approx_stat <- ((W^(-1 / d3) - 1) * (d1 * d3 - 2 * d2)) / (m  * h)

# The corresponding p-value is:
pf(F_approx_stat, m * h, d1 * d3 - 2 * d2, lower.tail = FALSE)
```

Compute a bootstrap distribution of the likelihood ratio test statistic for the test of the hypothesis $H_0:β_{1}=β_{2}=\ldots=β_{19}=0$ and compare it to the approximate $F$ distribution and to the approximate $χ^2$ distribution. Are the exact bootstrap distribution and its approximations matching well? What is the p-value for the test based on each of the three methods? 



```{r}
##rm(list = ls())

# Bootstrap study:
set.seed(697)
S <- 10####### 10000

LR_stat_sample <- numeric(S)
R1 <- numeric(S)
R2 <- numeric(S)
R3 <- numeric(S)

s <- 1

n<-nrow(df)
while (s <= S) {
  try({
    # resample - sample from the original data frame with replacements
    bootstrap_sample_idx <- sample(n, replace = TRUE)
    bstp_df <- df[bootstrap_sample_idx, ]

    
    res_full_model <- lm(cbind(FOODINSEC_15_17, CH_FOODINSEC_14_17, VLFOODSEC_15_17) ~ ., data = bstp_df)
    summary <- summary(res_full_model)
  
    R1[s] <- summary$`Response FOODINSEC_15_17`$r.squared
    R2[s] <- summary$`Response CH_FOODINSEC_14_17`$r.squared
    R3[s] <- summary$`Response VLFOODSEC_15_17`$r.squared
    
    res_partial_model <- lm(cbind(FOODINSEC_15_17, CH_FOODINSEC_14_17, VLFOODSEC_15_17) ~ PCT_LACCESS_POP15+PCT_LACCESS_HHNV15+PCT_LACCESS_SNAP15+PCT_LACCESS_CHILD15+PCT_NHWHITE10+ PCT_NHASIAN10+PCT_NHNA10+PCT_NHPI10+FSRPTH16+PCT_SNAP17+ PCT_OBESE_ADULTS17+PCT_NHBLACK10+PCT_HISP10+GROCPTH16,data = bstp_df)
    summary(res_partial_model)
  
    estSigma <- t(resid(res_full_model)) %*% resid(res_full_model) / n
    estSigma1 <- t(res_partial_model$residuals) %*% res_partial_model$residuals / n
    
    LR_stat_sample[s] <- det(estSigma) / det(estSigma1)
    
    s <- s + 1
  })
}

sum( LR_stat_sample < W ) / S

# CI for R^2
quantile(R1, c(0.025, 0.975)) # define CI for R^2 of FOODINSEC_15_17
quantile(R2, c(0.025, 0.975)) # define CI for R^2 of CH_FOODINSEC_14_17
quantile(R3, c(0.025, 0.975)) # define CI for R^2 of VLFOODSEC_15_17


h <- length(variables_to_test)
d1 <- (n - r - 1) - 0.5 * (m - h + 1)
d2 <- (m * h - 2)/4
d3 <- m^2 + h^2 - 5
d3 <- if (d3 > 0) { 
  sqrt(((m * h)^2 - 4) / d3)
} else { 1 }

F_approx_stat <- ((LR_stat_sample^(-1 / d3) - 1) * (d1 * d3 - 2 * d2)) / (m  * h)

crit_F <- ((W^(-1 / d3) - 1) * (d1 * d3 - 2 * d2)) / (m  * h)

hist(F_approx_stat, freq = FALSE, breaks = 40, main = "Bootstrap and F approximation")
rng <- range(F_approx_stat)
x <- seq(rng[1], rng[2], length.out = 100)
lines(x, df(x, m * h, d1 * d3 - 2 * d2))
abline(v = crit_F)
```

```{r}
# Rescale the bootstrap sample to the approximate F scale:
d1 <- (n - r - 1) - 0.5 * (m - h + 1)
d2 <- (m * h - 2)/4
d3 <- m^2 + h^2 - 5
d3 <- if (d3 > 0) { 
  sqrt(((m * h)^2 - 4) / d3)
} else { 1 }

F_approx_stat <- ((LR_stat_sample^(-1 / d3) - 1) * (d1 * d3 - 2 * d2)) / (m  * h)

hist(F_approx_stat, freq = FALSE, breaks = 40, main = "Bootstrap and F approximation")
rng <- range(F_approx_stat)
x <- seq(rng[1], rng[2], length.out = 100)
lines(x, df(x, m * h, d1 * d3 - 2 * d2))
```

```{r}
# Rescale the bootstrap sample to the approximate chi squared scale:
chi_sq_approx_stat <- -(n - r - 1 - 0.5 * (m - r + q + 1)) * log(LR_stat_sample)

hist(chi_sq_approx_stat, freq = FALSE, breaks = 40, main = "Bootstrap and chi-square approximation")
rng <- range(chi_sq_approx_stat)
x <- seq(rng[1], rng[2], length.out = 100)
lines(x, dchisq(x, m * (r - q)))
```